---
title: "An introduction to RiskMap"
vignette: >
  %\VignetteIndexEntry{An introduction to RiskMap}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

**RiskMap** contains functions for model based geostatistics (MBG).
MBG takes point based observations and generate predictions across the entire area of interest.
A comprehensive guide to using the package and MBG in general is available at [https://www.mbgr.org/] but this vignette provides a basic overview of the package's functionality.

## Installation and loading

Install **RiskMap** using:

```r
install.packages('RiskMap')
```

or from GitHub using:
  
```r
remotes::install_github("claudiofronterre/RiskMap")
```

Once installation is complete, load the package with:

```r
library(RiskMap)
```

## Datasets

The package contains 11 example datasets. 
Apart from `liberia`, all are provided in a format ready to analyse.
They can be loaded using `data(<dataset>)` e.g. `data(anopheles)`.
For more information on the datasets use `help(<dataset>)`

- `abund_sma` - Abundance of *Culex pipiens* mosquitos in the Sacramento Metropolitan Area
- `anopheles` - *Anopheles* mosquitoes in Southern Cameroon
- `galicia` -  Heavy metal biomonitoring in Galicia
- `infect_sma` - PCR-tested mosquito pools in the Sacramento Metropolitan Area
- `italy_sim` - Simulated dataset on the Italian peninsula
- `liberia` - Counts of riverblindness in Liberia
- `loaloa` - Prevalence of Loa loa (eyeworm) in West Africa
- `malkenya` - Surveys of malaria in Kenya
- `malnutrition` - Malnutrition in Ghana
- `tz_covariates` - Covariates for prediction of malaria in Tanzania
- `tz_malaria` - Malaria in Tanzania

## Loading data

Datasets need transforming into a simple features (sf) object prior to analysis using `st_as_sf()` from the sf package, supplying the names of the columns containing the spatial coordinates and the coordinate reference system for the coordinates:

```{r}
data(liberia)
sf_data <- sf::st_as_sf(liberia, coords = c("long", "lat"), crs = sf::st_crs(4326))

plot(sf_data["npos"])
```

For the other datasets, call `data()` and they are ready to use.

## Exploratory data analysis

```{r}

dist_summaries(sf_data, scale_to_km = TRUE)

plot_variogram(variogram(sf_data, 
                         variable = "npos", 
                         n_bins = 14, 
                         n_permutations = 100,
                         scale_to_km = TRUE), 
               plot_envelope = TRUE)


```




