---
title: "An introduction to RiskMap"
vignette: >
  %\VignetteIndexEntry{An introduction to RiskMap}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

**RiskMap** contains functions for model based geostatistics (MBG).
MBG takes point based observations and generate predictions across the entire area of interest.
A comprehensive guide to using the package and MBG in general is available at <https://www.mbgr.org/> but this vignette provides a basic overview of the package's functionality.

## Installation and loading

Install **RiskMap** using:

```r
install.packages('RiskMap')
```

or from GitHub using:
  
```r
remotes::install_github("claudiofronterre/RiskMap")
```

Once installation is complete, load the package with:

```r
library(RiskMap)
```

## Datasets

The package contains 11 example datasets. 
Apart from `liberia`, all are provided in a format ready to analyse.
They can be loaded using `data(<dataset>)` e.g. `data(anopheles)`.
For more information on the datasets use `help(<dataset>)`

- `abund_sma` - Abundance of *Culex pipiens* mosquitos in the Sacramento Metropolitan Area
- `anopheles` - *Anopheles* mosquitoes in Southern Cameroon
- `galicia` -  Heavy metal biomonitoring in Galicia
- `infect_sma` - PCR-tested mosquito pools in the Sacramento Metropolitan Area
- `italy_sim` - Simulated dataset on the Italian peninsula
- `liberia` - Counts of riverblindness in Liberia
- `loaloa` - Prevalence of Loa loa (eyeworm) in West Africa
- `malkenya` - Surveys of malaria in Kenya
- `malnutrition` - Malnutrition in Ghana
- `tz_covariates` - Covariates for prediction of malaria in Tanzania
- `tz_malaria` - Malaria in Tanzania

## Loading data

Datasets need transforming into a simple features (sf) object prior to analysis. With the exception of `liberia`, the datasets are already converted to this format. To convert a dataframe to a suitable format, use `st_as_sf()` from the sf package, supplying the names of the columns containing the spatial coordinates and the coordinate reference system for the coordinates:

```{r}
library(RiskMap)
data(liberia)
data <- sf::st_as_sf(liberia, coords = c("long", "lat"), crs = sf::st_crs(4326))

plot(data["npos"])
```

## Exploratory data analysis

The distances between data points can be summarised with `summarise_distance()`. By default, coordinates are converted to a projected coordinate reference system and reported in kilometers, but these can be overridden.

```{r}
summarise_distance(data)
```

The empirical variogram can be used to assess spatial independence in the data. In this example, `n_permutations` is set to 100 to speed up computation, but should be set to 10000 for final runs.

<need to explain succinctly what points outside the envelope signify>

```{r}
variogram <- variogram(data, 
                       variable = "npos", 
                       n_permutations = 100,
                       distance_units = "km")

plot_variogram(variogram, plot_envelope = TRUE)
```

## Model fitting

`glpgm()` fits generalised linear Gaussian process models. The model family can be either `gaussian`, `binomial` or `poisson`. For this dataset consisting of counts of disease incidence, a binomial model is most appropriate. To make the results reproducible, set a `seed` value in `set_control_mcmc()`.

```{r}
model <- glgpm(npos ~ log(elevation) + gp(),
               denominator = ntest, 
               data = data,
               model_crs = 32629,
               control_mcmc = set_control_mcmc(seed = 2026),
               return_samples = TRUE,
               family = "binomial",
               messages = FALSE)
```

We can access the model coefficients with `coef()`:

```{r}
coef(model)
```

Or summarise the model fit with `summary()`:

```{r}
summary(model)
```

For binomial and Poisson models, we can assess model convergence with `plot_mcmc()`:

```{r}
plot_mcmc(model)
```

## Generating predictions

To generate predictions from the model, we first need to generate locations where we wish to generate predictions for. Often, this involves obtaining boundary data for the region of interest, for which we can use the `geobounds` package and then generating a grid of points at a specified `spacing` inside it using `create_grid()`:

```{r}
library(geobounds)
liberia_boundary <- gb_get_adm0("Liberia")
liberia_boundary <- sf::st_transform(liberia_boundary, crs = 32629)
liberia_grid <- create_grid(liberia_boundary, spacing = 50)

plot(sf::st_geometry(liberia_boundary))
plot(liberia_grid, add = TRUE, col = 'red')
```

Alternatively, we can use `create_convex_hull()` to generate a boundary encompassing all of the locations in the data and generate a grid from this:

```{r}
liberia_hull <- create_convex_hull(data)
liberia_hull <- sf::st_transform(liberia_hull, crs = 32629)
liberia_hull_grid <- create_grid(liberia_hull, spacing = 50)

plot(sf::st_geometry(liberia_hull))
plot(liberia_hull_grid, add = TRUE, col = 'red')
```

After generating a grid, we need to generate a dataframe containing the variables included in the model for the new locations. In this example, we will obtain `elevation` from using the `elevatr` package:

```{r}
# Download elevation data
library(elevatr)
liberia_elev <- get_elev_raster(locations = liberia_grid, 
                                z = 6, 
                                clip = "locations",
                                verbose = FALSE)

# 3) Extracting elevation at the grid locations
new_data <- data.frame(elevation = 
                         terra::extract(liberia_elev,
                                sf::st_coordinates(liberia_grid)))

```

We can then use `setup_prediction()` to prepare the data to generate predictions either at the grid locations or for areal targets. To be able to generate predictions for areal targets, we need to set `type` to `"joint"`:

```{r}
prediction_prep <- setup_prediction(model, 
                                    grid_pred = liberia_grid,
                                    predictors = new_data,
                                    type = "joint",
                                    messages = FALSE)
```

```{r}
grid_prediction <- predict_grid_target(prediction_prep,
                                       f_target = list(prev = function(lp) exp(lp)/(1+exp(lp))))

plot(grid_prediction)
```



